---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-"
)
library(bbousims)
```

# bbousims

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![R-CMD-check](https://github.com/poissonconsulting/bbousims/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/poissonconsulting/bbousims/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/poissonconsulting/bbousims/branch/main/graph/badge.svg)](https://app.codecov.io/gh/poissonconsulting/bbousims?branch=main)
<!-- badges: end -->

`bbousims` is a package to simulate Boreal Caribou population abundance over time from survival, ageing and birth processes. 
Survival and recruitment data from hypothetical composition surveys and collaring are simulated from key sampling parameters.

The output of `bbs_simulate_caribou()` is intended to be used as input data for [bboutools](https://poissonconsulting.github.io/bboutools/) functions for fitting survival and recruitment models and predicting population growth. 

In addition, there is more general functionality for simulating population abundance given any set of stages, period duration, and survival, ageing and birth process matrices.

## Installation

You can install the development version of bbousims from [GitHub](https://github.com/) with:

``` r
# install.packages("remotes")
remotes::install_github("poissonconsulting/bbousims")
```

## Introduction

### Simulate survival and fecundity rates
Simulate survival rates for each month, year and stage and fecundity rates for each year and stage.
Rates are generated for female stages (female calf, female yearling and female adult).
Female calf and female adult survival rates are produced from separate linear models, whereas female yearling survival is determined as an effect on female adult survival. 
Variation in recruitment can be specified either through calf survival and/or calves per adult female. 

```{r}
set.seed(1)
survival <- bbs_survival_caribou(survival_adult_female = 0.85, 
                                 annual_sd_adult_female = 0.2,
                                 trend_adult_female = -0.1,
                                 month_sd_adult_female = 0.1,
                                 survival_calf_female = 0.5, 
                                 yearling_effect = 0.05,
                                 nyear = 5)

fecundity <- bbs_fecundity_caribou(calves_per_adult_female = 0.7, 
                                   annual_sd = 0.1,
                                   nyear = 5)
```
View expected monthly survival rates ('eSurvival') for adult females (stage 3) by each month and year.

```{r}
survival$eSurvival[,,3]
```

### Project population
Population is projected from survival and fecundity rates. 
Survival occurs at the end of each month and survival, ageing and birth occur at the end of each year, in that order.  

Initial population abundance for each stage is determined from the initial number of adult females set by the user and the calculated stable stage distribution (see `bbs_demographic_summary()` for details).
Population abundance for male stages are based on user-provided sex ratios. 


```{r}
set.seed(1)
population <- bbs_population_caribou(survival,
                                     fecundity = fecundity,
                                     adult_females = 500, 
                                     proportion_adult_female = 0.65)
```
The output is a matrix with abundance for each period and stage. The first period is the initial population and period 13 is the final month of the first year. 

```{r}
# projected population for first year
population[,1:13]
```

```{r, fig.height=4,fig.width=7}
bbs_plot_population(population)
```


### Simulate abundance, survival and recruitment data
Abundance, survival and recruitment data are simulated from hypothetical composition surveys and collaring, given the survival and fecundity rates used to project the population and a set of key sampling parameters. 
The output is a list of lists of the abundance, survival, and recruitment data.frames for each simulation.


```{r}
set.seed(1)
data <- bbs_simulate_caribou(survival = survival,
                             fecundity = fecundity,
                             nsims = 10,
                             adult_females = 500,
                             proportion_adult_female = 0.65,
                             month_composition = 9L,
                             collared_adult_females = 30,
                             group_size = 6, 
                             group_coverage = 0.3)
```

```{r, fig.height=4,fig.width=7}
bbs_plot_population(data)
```

View collar survival data for the first simulation
```{r}
# collar survival data for first simulation
print(data[[1]]$survival)
```

### Work with `bboutools`
The survival and recruitment data.frames in the output of `bbs_simulate_caribou()` are intended to be used as input data for model fitting functions in the [bboutools package]("https://poissonconsulting.github.io/bboutools/).

```{r, eval=FALSE}
# fit model for each simulation
# we set year_start = 1 because we assume the projected population is for the biological year
fits <- lapply(1:length(data), function(x){
  survival <- data[[x]]$survival
  bboutools::bb_fit_survival(data = survival, year_start = 1L)
})
```

## Information

Additional information is available from the [bbousims website](https://poissonconsulting.github.io/bbousims/), including more in-depth articles.

## bbou Suite

`bbousims` is part of the bbou suite of tools. 
Other packages in this suite include:

- [bboudata](https://github.com/poissonconsulting/bboudata)
- [bbouretro](https://github.com/poissonconsulting/bbouretro)
- [bboushiny](https://github.com/poissonconsulting/bboushiny)
- [bboutools](https://github.com/poissonconsulting/bboutools)

